Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-13 paper.qmd
Changed MSE equation to RMSE
Updated Figure 5

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = sqrt_row_wise_total_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

error_plot_two_curvy_hist <- 
  ggplot(error_df_two_curvy_abs) +
  geom_histogram(aes(x=sqrt_row_wise_total_error, y=..density..)) +
  geom_density(aes(x=sqrt_row_wise_total_error, y=..density..), colour="red") +
  #xlab(expression(group("|", e[hj], "|"))) +
  xlab(expression(e[hj])) +
  ylab("") +
  theme_bw() 
```

@@

The error is computed by taking the squared \pD{} Euclidean distance, corresponding to computing the mean squared error (MSE) as:

$$\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2$${#eq-equation1}

where $n$ is the number of observations, $b$ is the number of bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables, $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hj} = \sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2$ to be the residual for each observation.

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates square root of total error, dark colour indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the non-linear cluster with dense corner."
#| out-width: 100%
#| fig-pos: H

error_plot_two_curvy + error_plot_two_curvy_hist +
  plot_layout(ncol=2)
```

>>>>

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggbeeswarm)

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = sqrt_row_wise_total_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  xlab("tSNE1") +
  ylab("tSNE2") +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    legend.position = "none"
  )

# error_plot_two_curvy_hist <- 
#   ggplot(error_df_two_curvy_abs) +
#   geom_histogram(aes(x=sqrt_row_wise_total_error, y=..density..)) +
#   geom_density(aes(x=sqrt_row_wise_total_error, y=..density..), colour="red") +
#   #xlab(expression(group("|", e[hj], "|"))) +
#   xlab(expression(e[hj])) +
#   ylab("") +
#   theme_bw()

error_plot_two_curvy_quasi <- ggplot(
  error_df_two_curvy_abs, 
  aes(x = sqrt_row_wise_total_error, 
      y = 0, 
      colour = sqrt_row_wise_total_error)) +
  geom_quasirandom(alpha = 0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  xlab(expression(e[hj])) +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none")
```

@@

The error is computed by taking the squared \pD{} Euclidean distance, corresponding to computing the root mean squared error (RMSE) as:

$$\sqrt{\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$${#eq-equation1}

where $n$ is the number of observations, $b$ is the number of bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables, $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hj} = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$ to be the residual for each observation.

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates error ($e_{hj}$), dark colour indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the non-linear cluster with dense corner."
#| out-width: 100%
#| fig-pos: H

error_plot_two_curvy + error_plot_two_curvy_quasi +
  plot_layout(ncol=2)
```
